Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-02-24 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
updating sec 4.5, case sutdy I and II, and adding new secs 4.5.2 and 4.8

\author{Gabriel Riutort-Mayol \and Paul-Christian Bürkner \and Michael R.\ Andersen \and Arno~Solin \and Aki Vehtari
}

@@

A GP model with a periodic covariance function does no fit in the framework of the HSGP approximation covered in this study as a periodic covariance function has not a spectral representation, but it has also a low-rank representation. In Appendix~\ref{sec_periodic}, we briefly present the approximate linear representation of a periodic squared exponential covariance function as developed by \citet{solin2014explicit}, analyze the accuracy of this approximation and, finally, derive the GP model with this approximate periodic squared exponential covariance function.

@@

Ultimately, these recommendations are based on the relationships among the number of basis functions $m$, the boundary condition $L$, and the length-scale $\ell$, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the squared exponential and the Mat{\'e}rn ($\nu=3/2$ and $\nu=5/2$) covariance functions in the present section, and for the periodic squared exponential covariance function in Appendix~\ref{sec_periodic}. For other kernels, the relationships will be slightly different depending on the smoothness or wigglyness of the covariance function.

@@

Next, we analyze how the interaction between $m$ and $c$ affects the quality of the approximation. The length-scale and marginal variance of the covariance function will no longer be fixed but instead we compute the joint posterior distribution of the function values and the hyperparameters using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021} for both the exact GP and the HSGP models. Figure~\ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the model for varying $m$ and~$c$. Figure~\ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure~\ref{fig5_lscale_vs_J} shows the estimated length-scale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure~\ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computation time for this example would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$ could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure~\ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:
%
\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa). This is consistent with the expression for the eigenvalues in eq.~\eqref{eq_eigenvalue}, where $L$ appears in the denominator.
\item There exists a minimum $c$ below which an accurate approximation will never be achieved regardless of the number of basis functions $m$.
\end{itemize}

\subsection{Near linear proportionality between $m$, $c$ and $\ell$} \label{subsec_theoretic_linearity}

@@

When considering squared exponential and Mat{\'e}rn covariance functions, we can show with simple algebra that when when $c$ is larger than the minimal value recommendation, the number of $m$ first terms needed to explain almost 100\% of the variation has a near linear relationship with $\ell$ and $c$. With decreasing $\ell$, the $m$ should grow near linearly with $1/\ell$, and with increasing $c$, the $m$ should grow near linearly with $c$.
%
This is natural as with decreasing $\ell$, more higher frequency basis functions are needed. With increasing $c$, as a smaller range of the basis functions are used in the approximation, the expected number of zero up-crossings goes down linearly with $c$, and thus more higher frequency basis functions are needed to compensate this.
%
When $c$ is below our recommendations given $\ell$, the effect of $\ell$ and $c$ to the recommended $m$ is more non-linear, but as long as we stay in the recommended range the linearity assumption is useful thumb rule for how to change $m$, if $\ell$ or $c$ are changed.

% There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the length-scale $l$.
% In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

% {\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
% proportional to 1/l which gives as that part, and the effect of c should also
% be quite linear when c is not very close to 1 and m is not very small as it
% linearly makes the interval shorter so that the there are less zero-level
% upcrossings available in the basis functions. You can visualize this using
% {\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
% look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
% (ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
% spectral densities and when they go to zero. This is illustrated in the same
% notebook after that figure. E.g. with length-scale l=1,l=1/2, l=1/4, l=1/8 to
% explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
% (linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
% basis functions. These spectral densities don't depend on c. When c increaes
% the basis functions are truncated and for example going from c=1.5 to c=2.25,
% basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
% functions are needed. So we have proportionality c/l. }

@@

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig6_relationships.png}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), for the squared exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions.}
  \label{fig6_relationships}
\end{figure*}

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $l$. Figure~\ref{fig6_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with squared exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure~\ref{fig6_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq.~\eqref{eq_diff_covs}. We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq.~\eqref{eq_diff_covs}, is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:
%
\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figure~\ref{fig6_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

\subsubsection{Relationships between $m$ and $\ell$ for a periodic squared exponential covariance function} \label{subsec_empiric_relations_periodic}

In Appendix~\ref{sec_periodic}, we analyze the accuracy of the linear representation of a periodic squared exponential covariance function and derive the minimum number of terms $m$ in the approximation required to achieve a close approximation to the exact periodic squared exponential kernel as a function of the length-scale $\ell$ of the kernel. Since this is a series expansion of sinusoidal functions, the approximation does not depend on any boundary condition. This relationship between $m$ and $\ell$ is gathered in Figure~\ref{figB1_m_lscale_periodic}.

@@

As explained in Section~\ref{subsec_theoretic_linearity}, when $c$ is large enough, there is a near linear proportionality between $m$, $\frac{\ell}{S}$ (length-scale $\ell$ relative to the half range of the input domain $S$) and $c$. To obtain practical numerical functions that can be used to guide the selection of these parameters, we have empirically derived the practically useful constant terms. We require a lower bound for $c$ of $c \geq 1.2$ such that the equations below are precise enough for practical application.

@@

\paragraph*{Squared exponential}:
%
\begin{align}\label{eq_m_l_QE}
&m= 1.75 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 1.75 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_QE}
&c \geq \, 3.2 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2 .
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}\label{eq_m_l_mat52}
&m= 2.65 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 2.65 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat52}
&c \, \geq \, 4.1 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2 .
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}\label{eq_m_l_mat32}
&m= 3.42 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 3.42 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat32}
&c \, \geq \, 4.5 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2 .       
\end{align}

These constants vary monotonically with respect to $\nu$ (squared exponential corresponding to Mat\'ern with $\nu \to \infty$). Using the formula for Mat\'ern ($\nu$=3/2) provides the largest $m$ and $c$, and thus this formula alone could be used as a conservative choice for all Mat\'ern covariance functions with $\nu \geq 3/2$ and likely as a good initial guess for many other covariance functions. If the aim is to find minimal $m$ to speedup the computation, a further refined formula can be obtained for new covariance functions.

Figure~\ref{fig6_relationships} and previous equations~(\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) were obtained for a GP with a unidimensional covariance function, which result in a surfaces depending on three variables, $m$, $c$ and $\ell$. Equivalent results for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figure~\ref{fig6_relationships} or equations~(\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) to check the accuracy by analyzing individually the different dimensions of a multidimensional GP model.

\subsubsection{Numerical equation for a periodic squared exponential covariance function} \label{sec_num_equation_periodic}

The numerical equation governing the relationships between $m$ and $\ell$ for a periodic squared exponential covariance function was estimated in eq.~\eqref{eq_J_l_periodic} which is depicted next:
%
\begin{align*}
&m \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{m} .
\end{align*}

@@

As stated above, equations (\ref{eq_m_l_QE}), (\ref{eq_m_l_mat32}) and (\ref{eq_m_l_mat52}) (depending on which kernel is used) provide the minimum length-scale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the length-scale $\ell$ controls the wigglyness of the function, it strongly influences the difficulty of obtaining accurate inference about the function from the data. Basically, if the length-scale estimate is accurate, we can expect the HSGP approximation to be accurate as well.

Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum length-scale provided by equations (\ref{eq_m_l_QE}), (\ref{eq_m_l_mat32}) and (\ref{eq_m_l_mat52}) (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum length-scale, the approximation is assumed to be good. If, however, $\hat{\ell}$ does not exceed the recommended minimum length-scale, the approximation may be inaccurate and $m$ and/or $c$ need to be updated. If we look back to the conclusions drawn from Figures~\ref{fig4_MSE_vs_J} and \ref{fig5_lscale_vs_J}, where $m = 10$ basis functions and a boundary factor of $c = 1.5$ were sufficient to obtain an accurate approximation of a function with $\ell = 0.3$, we can recognize that these conclusions also match those obtained from eq. (\ref{eq_m_l_QE}).

@@

Guidelines for updating $m$ and/or $c$ are:
%
\begin{itemize}
	\item $c$ must be big enough for the given length-scale $l$ of the function to be learned and,
	\item $m$ must be big enough for the given $\ell$ and $c$. 
\end{itemize}

We recommend using this diagnostic in an iterative procedure by starting from some initial guess of $\ell$ and obtain the optimum valid values for $c$ and $m$ as a function of this initial guess of $\ell$ by using the equations in Section~\ref{sec_num_equations}; and if the estimated $\hat{\ell}$ is below the minimum length-scale determined by $c$ and $m$, update $\ell$ with $\hat{\ell}$ and repeat the process.

If the initial length-scale is not far from the optimal, and larger than minimal $c$ and $m$ are used in the initial HSGP model, it is likely that the initial results are already sufficiently accurate. The following iterative procedure works as safety approach and also allows finding a smaller sufficient $m$ which can be useful if the user wants to optimize the computational speed for repeated fitting (e.g., in cross-validation or simulation based calibration).

%A detailed guideline with the steps to perform diagnostic, ensuring that a good approximation is achieved with the minimum computational requirements (i.e. minimum number of basis functions $m$) is provided in  next Section~\ref{sec_user_guide}.

@@

Based on the above proposed diagnostics, we obtain a simple, iterative step-by-step procedure that users can follow in order to obtain an accurate HSGP approximation. The procedure is split into two phases, Phase~A and B, which have to be completed consecutively.
%In the following, we will refer to the Equations from Section XXX for the double-exponential kernel, but the same procedure works for other kernels by using the corresponding Equations provided in the same Section.

@@

\item[A1.] Make an initial guess on the length-scale $\ell_1$. If there is no useful information available, we recommend to start with a large length-scale, about $\ell_1 \in [0.5, 1]$. Set the iteration index to $i=1$.

\item[A2.] Obtain the minimum valid boundary factor $c_i$ determined by $\ell = \ell_i$ for the given kernel as per Section~\ref{sec_num_equations}.

\item[A3.] Obtain the mimimum valid number of basis functions $m_i$ determined by $\ell = \ell_i$ and $c = c_i$ for the given kernel as per Section~\ref{sec_num_equations}.

@@

\item[A5.] Perform the length-scale diagnostic by checking if $\hat{\ell}_i - 0.01 \geq \ell_i$.

@@

If the diagnostic is TRUE, the HSGP approximation is close to be sufficiently accurate. Continue with Phase~B.

@@

\item[B1.] For the current HSGP model, compute measures of predictive accuracy, for example, root mean square error (RMSE), coefficient of determination ($R^2$), or expected log predictive density (ELPD).

@@

\item[B4.] Obtain the minimum valid length-scale $\ell_i$ that can be accurately estimated by $m = m_i$ and $c = c_i$ for the given kernel as per Section~\ref{sec_num_equations}.

@@

\item[B6.] Perform the length-scale diagnostic by checking if $\hat{\ell}_i - 0.01 \geq \ell_i$.

@@

If all the stability checks succeed, the HSGP approximation of the latest model should be sufficiently accurate and the procedure ends here.

@@

>>>>

\author{Gabriel Riutort-Mayol \and Paul-Christian Bürkner \and Michael R.\ Andersen \and Arno Solin \and Aki Vehtari
}

@@

A GP model with a periodic covariance function does not fit in the framework of the HSGP approximation covered in this study as a periodic covariance function has not a spectral representation, but it has also a low-rank representation. In Appendix~\ref{sec_periodic}, we briefly present the approximate linear representation of a periodic squared exponential covariance function as developed by \citet{solin2014explicit}, analyze the accuracy of this approximation and, finally, derive the GP model with this approximate periodic square exponential covariance function.

@@

Ultimately, these recommendations are based on the relationships among the number of basis functions $m$, the boundary condition $L$, and the length-scale $\ell$, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the square exponential and the Mat{\'e}rn ($\nu=3/2$ and $\nu=5/2$) covariance functions in the present section, and for the periodic squared exponential covariance function in Appendix~\ref{sec_periodic}. For other kernels, the relationships will be slightly different depending on the smoothness or wigglyness of the covariance function.

@@

Next, we analyze how the interaction between $m$ and $c$ affects the quality of the approximation. The length-scale and marginal variance of the covariance function will no longer be fixed but instead we compute the joint posterior distribution of the function values and the hyperparameters using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021} for both the exact GP and the HSGP models. Figure~\ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the model for varying $m$ and $c$. Figure \ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure \ref{fig5_lscale_vs_J} shows the estimated length-scale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computation time for this example would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$  could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure~\ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:

\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa). This is consistent with the expression for the eigenvalues in eq.~\eqref{eq_eigenvalue}, where $L$ appears in the denominator.
\item There exists a minimum $c$ below which an accurate approximation will never be achieved regardless of the number of basis functions $m$.
\end{itemize}

\subsection{Near linear proportionality between $m$, $c$, and $\ell$} \label{subsec_theoretic_linearity}

@@

When considering squared exponential and Mat{\'e}rn covariance functions, we can show with simple algebra that when $c$ is larger than the minimal value recommendation, the number of $m$ first terms needed to explain almost 100\% of the variation has a near linear relationship with $\ell$ and $c$. With decreasing $\ell$, the $m$ should grow near linearly with $1/\ell$, and with increasing $c$, the $m$ should grow near linearly with $c$.
%
This is natural as with decreasing $\ell$, more higher frequency basis functions are needed. With increasing $c$, as a smaller range of the basis functions are used in the approximation, the expected number of zero up-crossings goes down linearly with $c$, and thus more higher frequency basis functions are needed to compensate this.
%
When $c$ is below our recommendations given $\ell$, the effect of $\ell$ and $c$ to the recommended $m$ is more non-linear, but as long as we stay in the recommended range the linearity assumption is useful thumb rule for how to change $m$, if $\ell$ or $c$ are changed.

% There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$.
% In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

% {\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
% proportional to 1/l which gives as that part, and the effect of c should also
% be quite linear when c is not very close to 1 and m is not very small as it
% linearly makes the interval shorter so that the there are less zero-level
% upcrossings available in the basis functions. You can visualize this using
% {\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
% look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
% (ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
% spectral densities and when they go to zero. This is illustrated in the same
% notebook after that figure. E.g. with lengthscale l=1,l=1/2, l=1/4, l=1/8 to
% explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
% (linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
% basis functions. These spectral densities don't depend on c. When c increaes
% the basis functions are truncated and for example going from c=1.5 to c=2.25,
% basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
% functions are needed. So we have proportionality c/l. }

@@

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig6_relationships.png}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), for the square exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions.}
  \label{fig6_relationships}
\end{figure*}

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $\frac{\ell}{S}$ (lengthscale $\ell$ relative to the half-range $S$ of the input domain). Figure \ref{fig6_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with squared exponential, Mat\'ern ($\nu$=3/2), and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with length-scale $\frac{\ell}{S}$ (From now to later on we will refer to $\frac{\ell}{S}$ by just $\ell$) and given a boundary factor $c$, Figure \ref{fig6_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which an accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figure~\ref{fig6_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

As explained in Section~\ref{subsec_theoretic_linearity}, when $c$ is large enough, there is a near linear proportionality between $m$, $\ell$ and $c$. To obtain practical numerical functions that can be used to guide the selection of these parameters, we have empirically derived the practically useful constant terms. We require a lower bound for $c$ of $c \geq 1.2$ such that the equations below are precise enough for practical application.

@@

\paragraph*{Squared exponential}:
%
\begin{align}
&m= 1.75 \, \frac{c}{\nicefrac{\ell}{S}} \;\; \Leftrightarrow \;\; \nicefrac{\ell}{S}= 1.75 \, \frac{c}{m},	\label{eq_m_l_QE} \\
&\text{with} \nonumber \\
&c \geq \, 3.2 \, \nicefrac{\ell}{S} \;\; \& \;\, c \geq \, 1.2	\label{eq_c_vs_l_QE}
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}
&m= 2.65 \, \frac{c}{\nicefrac{\ell}{S}} \;\; \Leftrightarrow \;\; \nicefrac{\ell}{S}= 2.65 \, \frac{c}{m}, \label{eq_m_l_mat52} \\
&\text{with} \nonumber \\
&c \, \geq \, 4.1 \, \nicefrac{\ell}{S} \;\; \& \;\, c \geq \, 1.2  \label{eq_c_vs_l_mat52}
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}
&m= 3.42 \, \frac{c}{\nicefrac{\ell}{S}} \;\; \Leftrightarrow \;\; \nicefrac{\ell}{S}= 3.42 \, \frac{c}{m}, \label{eq_m_l_mat32}\\
&\text{with} \nonumber \\
&c \, \geq \, 4.5 \, \nicefrac{\ell}{S} \;\; \& \;\, c \geq \, 1.2 \label{eq_c_vs_l_mat32}
\end{align}
%
These constants vary monotonically with respect to $\nu$ (squared exponential corresponding to Mat\'ern with $\nu \rightarrow \infty$). Using the formula for Mat\'ern ($\nu$=3/2) provides the largest $m$ and $c$, and thus this formula alone could be used as a conservative choice for all Mat\'ern covariance functions with $\nu \geq 3/2$ and likely as a good initial guess for many other covariance functions. If the aim is to find minimal $m$ to speedup the computation, a further refined formula can be obtained for new covariance functions.

Figure~\ref{fig6_relationships} and previous equations (\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) were obtained for a GP with a unidimensional covariance function, which result in a surfaces depending on three variables, $m$, $c$ and $\ell$. Equivalent results for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figure \ref{fig6_relationships} or equations (\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) to check the accuracy by analyzing individually the different dimensions of a multidimensional GP model.

\subsection{Relationships between $m$ and $\ell$ for a periodic squared exponential covariance function} \label{subsec_empiric_relations_periodic}

As commented in Section~\ref{sec_method_periodic}, in Appendix~\ref{sec_periodic} we present an approximate linear representation of a periodic squared exponential covariance function. In Appendix~\ref{sec_periodic}, we also analyze the accuracy of this linear representation and derive the minimum number of terms $m$ in the approximation required to achieve a close approximation to the exact periodic squared exponential kernel as a function of the length-scale $\ell$ of the kernel. Since this is a series expansion of sinusoidal functions, the approximation does not depend on any boundary condition. This relationship between $m$ and $\ell$ for a periodic squared exponential covariance function is gathered in Figure~\ref{figB1_m_lscale_periodic} and the numerical equation was estimated in eq.~\eqref{eq_J_l_periodic} which is depicted next:
%
\begin{align*}
&m \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{m}
\end{align*}

@@

As stated above, equations \eqref{eq_m_l_QE}, \eqref{eq_m_l_mat52}, \eqref{eq_m_l_mat32} and \eqref{eq_J_l_periodic} (depending on which kernel is used) provide the minimum length-scale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the length-scale $\ell$ controls the wigglyness of the function, it strongly influences the difficulty of obtaining accurate inference about the function from the data. Basically, if the length-scale estimate is accurate, we can expect the HSGP approximation to be accurate as well.

Having obtained an estimate $\hat{\ell}$ of a given $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum length-scale provided as a function of $m$ and $c$ by equations \eqref{eq_m_l_QE}, \eqref{eq_m_l_mat52}, \eqref{eq_m_l_mat32} and \eqref{eq_J_l_periodic} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum length-scale, the approximation is assumed to be good, or at least close to. If, however, $\hat{\ell}$ does not exceed the recommended minimum length-scale, the approximation may be inaccurate and $m$ and/or $c$ need to be updated. If we look back to the conclusions drawn from Figures~\ref{fig4_MSE_vs_J} and \ref{fig5_lscale_vs_J}, where $m = 10$ basis functions and a boundary factor of $c = 1.5$ were sufficient to obtain an accurate approximation of a function with $\ell = 0.3$, we can recognize that these conclusions also match those obtained from equations \eqref{eq_m_l_QE} and \eqref{eq_c_vs_l_QE}.

Equations in Section~\ref{sec_num_equations} to update $m$ and $c$ imply:
% 
\begin{itemize}
	\item $c$ must be big enough for the given length-scale $\ell$ of the function to be learned and,
	\item $m$ must be big enough for the given $\ell$ and $c$. 
\end{itemize}
%
If larger than minimal $c$ and $m$ for the given length-scale $\ell$ are used in the initial HSGP model, it is likely that the initial results are already sufficiently accurate. However, as the given length-scale $\ell$ of the function to be learned is unknown, we recommend using this diagnostic in an iterative procedure by starting from some initial guess of $\ell$ and obtain the optimum valid values for $c$ and $m$ as a function of this initial guess of $\ell$ by using the equations in Section~\ref{sec_num_equations}; and if the estimated $\hat{\ell}$ is below the minimum length-scale determined by $c$ and $m$, which indicates a potential problem with the approximation, update $\ell$ with $\hat{\ell}$ and repeat the process until both:
%
\begin{itemize}
\item the estimated $\hat{\ell}$ is larger than the minimum lengthscale determined by $c$ and $m$, which indicates the approximation must be sufficiently accurate, and
\item the measures of predictive accuracy, for example, root mean square error (RMSE), coefficient of determination ($R^2$), or expected log predicitve density (ELPD) reach a stabilization.
\end{itemize}

As commented in next Section~\ref{sec_diagnostic_criticism}, the estimated $\hat{\ell}$ being larger than the minimum lengthscale determined by $c$ and $m$ is not a definite certain that the approximate is sufficiently accurate, although it is for most of the practical cases. Therefore, we recommend to look at the residual error of the fit until it stabilizes.

@@

%If the initial lengthscale is not far from the optimal, and larger than minimal $c$ and $m$ are used in the initial HSGP model, it is likely that the initial results are already sufficiently accurate. 
The iterative procedure, briefly commented above and detailed in next Section~\ref{sec_user_guide}, works as safety approach and also allows finding a smaller sufficient $m$ which can be useful if the user wants to optimize the computational speed. Using the optimum values for $m$ and $c$ that allow for minimum computational requirements can be essential for repeated fitting (e.g., in cross-validation or simulation based calibration), uni-dimensional functions with very large data sets, and even more essential in multi-dimensional cases ($D \leq 4$; notice from Section~\ref{sec_method_multi} that for $D > 4$ HSGPs start to be unpractical in any case) where knowing the optimum values for $m$ and $c$ for each individual dimension can significantly reduce the computational requirements in these multi-dimensional cases; otherwise fitting HSGPs would be unfeasible here.

@@

Based on the above proposed diagnostics, we obtain a simple, iterative step-by-step procedure that users can follow in order to obtain an accurate HSGP approximation. The procedure is split into two phases, Phase A and B, which have to be completed consecutively.
%In the following, we will refer to the Equations from Section XXX for the double-exponential kernel, but the same procedure works for other kernels by using the corresponding Equations provided in the same Section.

@@

\item[A1.] Make an initial guess on the lengthscale $\ell_1$. If there is no useful information available, we recommend to start with a large lengthscale, about $\ell_1 \in [0.5, 1]$. Set the iteration index to $i=1$.

\item[A2.] Obtain the minimum valid boundary factor $c_i$ determined by $\ell = \ell_i$ for the given kernel as per Section \ref{sec_num_equations}.

\item[A3.] Obtain the mimimum valid number of basis functions $m_i$ determined by $\ell = \ell_i$ and $c = c_i$ for the given kernel as per Section \ref{sec_num_equations}.

@@

\item[A5.] Perform the lengthscale diagnostic by checking if $\hat{\ell}_i - 0.01 \geq \ell_i$.

@@

If the diagnostic is TRUE, the HSGP approximation is close to be sufficiently accurate. Continue with Phase B.

@@

\item[B1.] For the current HSGP model, compute measures of predictive accuracy, for example, RMSE, $R^2$, or ELPD.

@@

\item[B4.] Obtain the mimimum valid lengthscale $\ell_i$ that can be accurately estimated by $m = m_i$ and $c = c_i$ for the given kernel as per Section \ref{sec_num_equations}.

@@

\item[B6.] Perform the lengthscale diagnostic by checking if $\hat{\ell}_i - 0.01 \geq \ell_i$.

@@

If all the stability checks succeeed, the HSGP approximation of the latest model should be sufficiently accurate and the procedure ends here.

@@

\subsubsection{"Criticism" of the diagnostic} \label{sec_diagnostic_criticism}

The basis of our diagnostic tool are the following:
% 
\begin{enumerate}
\item observing a length-scale estimate $\hat{\ell}$ smaller than recommended minimum length-scale determined by $m$ and $c$ indicates a potential problem with the approximation,
\item observing a $\hat{\ell}$ larger than the recommended minimum length scale implies that the approximation is accurate.
\end{enumerate}

The later statement, however, cannot be assumed to be true in general, that is, cannot be generalized to all possible cases. ({\color{red} TODO: A major revision from a reviewer was related to that, i.e. to be critical with the diagnostic and discuss about possible cases where it could fail. I don't know if we could discuss a bit about this. I remember Michael commented an example of a potential case where that diagnostic statement could not be true.})
